---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PULS

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/vinhtantran/puls.svg?branch=master)](https://travis-ci.com/vinhtantran/puls)
[![R-CMD-check](https://github.com/vinhtantran/puls/workflows/R-CMD-check/badge.svg)](https://github.com/vinhtantran/puls/actions)
<!-- badges: end -->

Partitioning using local subregions (PULS) is a clustering technique designed to explore subregions of functional data for information to split the curves into clusters.

## Installation

You can install the released version of `puls` from [CRAN](https://CRAN.R-project.org) with:

```r
install.packages("puls")
```

And the development version from [GitHub](https://github.com/) with:

```r
# install.packages("remotes")
remotes::install_github("vinhtantran/puls")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(puls)
library(fda)
# Build a simple fd object from already smoothed smoothed_arctic
data(smoothed_arctic)
NBASIS <- 300
NORDER <- 4
y <- t(as.matrix(smoothed_arctic[, -1]))
splinebasis <- create.bspline.basis(rangeval = c(1, 365),
                                    nbasis = NBASIS,
                                    norder = NORDER)
fdParobj <- fdPar(fdobj = splinebasis,
                  Lfdobj = 2,
                  # No need for any more smoothing
                  lambda = .000001)
yfd <- smooth.basis(argvals = 1:365, y = y, fdParobj = fdParobj)

Jan <- c(1, 31); Feb <- c(31, 59); Mar <- c(59, 90)
Apr <- c(90, 120); May <- c(120, 151); Jun <- c(151, 181)
Jul <- c(181, 212); Aug <- c(212, 243); Sep <- c(243, 273)
Oct <- c(273, 304); Nov <- c(304, 334); Dec <- c(334, 365)

intervals <-
  rbind(Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec)

PULS4_pam <- PULS(toclust.fd = yfd$fd, intervals = intervals,
                  nclusters = 4, method = "pam")
PULS4_pam
```

Output plot

You can also embed plots, for example:

```{r pressure, echo = FALSE}
# plot(PULS4_pam)
```
